---
title: "F1Modeling"
author: "Derek Rogers"
date: "2023-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
```

```{r}
F1DataFrame = read.csv("C:/SMU MS Data Science/AppliedStatistics6372/Project1/F1ParametricModeling/ModelingData/F1ModelingCustomCategoricalVariables.csv")

F1DataFrame
```

# Residuals Vs Fitted Plot 

Shows evidence of a negative slope with variance that seems to increase as fitted values increase. 

This means the model is on average predicting worse results for drivers who finish with a good results. As the fitted values increase it seems to get a wider difference in predicted vs actual results.


# Q-Q Plot

 suggests a normal distribution can be assumed
 
# Scale-Location Plot

 Provides more evidence for normality
 
 There seems to be a patterns in starting at each discrete fitted value and traveling parallel to the Y axis with a sin function

```{r}
mymodel<-lm(positionOrder ~ grid + name + fullname + era, data = F1DataFrame)
par(mfrow = c(2,2))
plot(mymodel)
par(mfrow = c(1,1))

summary(mymodel)
```
